---
title: "05NSLAngerAndrea"
author: "Andrea Anger"
date: "2025-03-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**ITDS 1372 Computer Programming Date: March 5, 2025
Lab 3: Conditions, loops, functions**


1. Basic if-else Statement
Instructions:
• Assign a value to a variable x.
• Check if x is greater than 5 using an if condition.
• Print "x is greater than 5" if true; otherwise, print "x is 5 or less".

```{r}
x<-4
if(x>5){
  print("x is greater than 5")
}
print("x is 5 or less")
```

2. if-else with Multiple Conditions
Instructions:
• Assign a value to x.
• Use if to check if x is greater than 0 and print "Positive number".
• Use else if to check if x is less than 0 and print "Negative number".
• If neither condition is true, print "Zero".

```{r}
x <-0
if(x>0){
  print("Positive number")
} else if (x<0){
  print("Negative number")
} else {
  print("Zero")
}
```

3. if-else with Logical Operators
Instructions:
• Assign an age value to age.
• Use if with & (AND) to check if age is between 18 and 60.
• If the condition is true, print "Eligible for work", else print "Not eligible for
work".

```{r}
age <- 17
if (age>=18 && age<=60){
  print ("Eligible for work")
} else {
  print("Not eligible for work")
}
```

4. if-else Inside a Function
Instructions:
• Define a function check_even_odd(num).
• Inside the function, use if to check if num is divisible by 2 (num %% 2 == 0).
• If true, return "Even number"; otherwise, return "Odd number".
• Call the function with a sample number and print the result.
ITDS 1372 Computer Programming Date: March 5, 2025

```{r}
check_even_odd <- function(num) {
  if (num %% 2 ==0){
    return("Even number")
  } else return("Odd number")
}
print(check_even_odd(-12))

```

5. Nested if-else
Instructions:
• Assign a score value.
• Use if to check if score is greater than or equal to 90 and print "Grade: A".
• If not, use another if inside else to check if score >= 75 and print "Grade: B".
• If neither condition is met, print "Grade: C".

```{r}
score<- 100
if (score>=90){
  print("Grade: A")
} else {
  if (score>=75) {
    print("Grade: B")
  } else {
    print("Grade: C")
  }
}
```

6. Using ifelse() for Vectorized Operations
Instructions:
• Create a vector numbers with multiple values.
• Use ifelse(condition, value_if_true, value_if_false) to check if each number
is even or odd.
• Store the results in a variable and print them.

```{r}
numbers <- c(1007, 95, 80, 76, 43)
results <- ifelse (test = numbers %% 2 ==0, yes= "Even", no="Odd")
print(results)
```

7. Basic switch Statement
Instructions:
• Assign a string value to day (e.g., "Monday").
• Use switch() with different cases ("Monday", "Friday", "Sunday", etc.).
• Print a corresponding message for each case.
• If the day doesn’t match any case, print "Regular day".

```{r}
day <- "Friday"
which_day <- switch(EXPR=day, 
           Monday=print("Monday"),
           Tuesday=print("Tuesday"),
           Wednesday=print("Wednesday"),
           Thursday=print("Thursday"),
           Friday=print("Friday"),
           print("Regular day")
         )
```

8. switch Statement Returning a Value
Instructions:
• Assign a fruit name ("apple", "banana", "grape", etc.) to fruit.
• Use switch() to return a color based on the fruit name.
• Store the result in a variable and print it.
ITDS 1372 Computer Programming Date: March 5, 2025

```{r}
fruit <- "banana"
which_color <- switch(EXPR=fruit, 
           apple="Red",
           grape="Purple",
           banana="Yellow",
           NA
         )
print(which_color)
```

9. switch with Numeric Cases
Instructions:
• Assign a numeric option (e.g., 2) to option.
• Use switch() with different cases to return "First option selected", "Second
option selected", etc.
• Print the result.

```{r}
option <- 3
which_option <- switch(EXPR=option, 
           '1'="First option selected",
           '2'="Second option selected",
           '3'="Third option selected",
           NA
         )
print(which_option)
```

10. switch in a Function
Instructions:
• Define a function operation(choice, a, b).
• Inside the function, use switch() to perform different operations:
o "add" → return a + b
o "subtract" → return a - b
o "multiply" → return a * b
o "divide" → check if b != 0, return a / b, else return "Cannot divide by
zero".
o Default case → return "Invalid choice".
• Call the function with different operations and print the results.

```{r}
operation <- function(choice,a,b){
  choice <- switch (EXPR = choice,
                    "add"=a+b,
                    "subtract"=a-b,
                    "multiply"=a*b,
                    "divide"=if (b !=0){
                      a/b
                    } else { "Cannot divide by zero"},
                    "Invalid choice"
  )
  return(choice)
}
print(operation("divide",3,0))
```

For Loop Problems
1. Print Numbers from 1 to 10
• Use a for loop to iterate from 1 to 10.
• Print each number.

```{r}
for(numbers in 1:10) {
  print(numbers)
}
```

2. Print Even Numbers from 2 to 20
• Use a for loop to iterate through even numbers from 2 to 20.
• Print each even number.

```{r}
for (even_number in 2:20) {
    if (even_number %% 2 == 0) {  # Check if even_number is even
        print(even_number)
    }
}
```

3. Calculate the Sum of First 10 Natural Numbers
ITDS 1372 Computer Programming Date: March 5, 2025
• Initialize a variable sum = 0.
• Use a for loop to add numbers from 1 to 10 to sum.
• Print the final sum.

```{r}
sum <- 0
for (num in 1:10){
  sum=sum+num
}
print(sum)
```

4. Print Elements of a Vector
• Create a vector v <- c(10, 20, 30, 40, 50).
• Use a for loop to print each element.

```{r}
v <- c(10,20,30,40,50)
for (i in 1:length(v) ){
  print(v[i])
}
```

5. Factorial of a Number Using for Loop
• Initialize fact = 1.
• Use a for loop to multiply numbers from 1 to n.
• Print the factorial value.

```{r}
fact <- 1
n <-4
for(num in 1:n){
  fact=fact*num
}
print(fact)
```

While Loop Problems
6. Print Numbers from 1 to 5 Using while Loop
• Initialize x = 1.
• Use a while loop to print numbers from 1 to 5.

```{r}
x <-1
while(x<=5){
  print(x)
  x=x+1
}
```

7. Find the First Number Divisible by 7 and 5
• Initialize num = 1.
• Use a while loop to check if num is divisible by 7 and 5.
• Stop the loop once the first such number is found.

```{r}
num <-1
while(num%%7!=0 | num%%5!=0){
  num=num+1
}
print(num)
```

8. Sum of Digits of a Number Using while Loop
• Given a number n = 1234, initialize sum = 0.
• Use a while loop to extract each digit and add it to sum.
• Print the final sum.

```{r}
sum <- 0
n <- 12345
i <-1
while(i<=nchar(as.character(abs(n)))){
  number_as_string <- as.character(abs(n))
  digit_to_add <- as.numeric(substr(number_as_string,i,i))
  sum=digit_to_add+sum
  i=i+1
}
print(sum)
```

ITDS 1372 Computer Programming Date: March 5, 2025
Repeat Loop Problems
9. Print "Hello" 5 Times Using repeat Loop
• Use a repeat loop to print "Hello" 5 times.
• Use counter variable to track iterations and use break to stop.

```{r}
x <- 1
repeat {
    print("Hello")
    x=x+1
    if (x > 5) {
        break
    } 
}
```

10. Find Square Root Using repeat Loop
• Use repeat to approximate the square root of n = 25 using Newton’s method.
• Stop when the difference between two successive approximations is very small.

```{r}
n <- 25
repeat {
    n<-sqrt(n)
    if (abs(sqrt(n)-n)<1e-10 ) {
        break
    } 
}
print(n)
```

Nested Loop Problems
1. Print a 5x5 Star Pattern
• Use a nested for loop.
• Print a * pattern of size 5x5.

```{r}
for(i in 1:5){
  stars<-""
  for(j in 1:5){
    stars <- paste(stars,"*",sep="")
  }
  cat(stars,"\n")
}
```

2. Multiplication Table (1 to 5)
• Use a nested for loop.
• Print the multiplication table for numbers 1 to 5.

```{r}
m<-matrix(0,5,5,byrow=FALSE,dimnames=list(c(1:5),c(1:5)))
for(i in 1:5){
  product<-0
  for(j in 1:5){
    product <- j*i
    m[i,j] = product
  }
}
print(m)
```

3. Print a Right-Angled Triangle Pattern
• Use a nested for loop.
• Print a right-angled triangle using numbers, where each row contains that row number.
Example Output:
1
2 2
3 3 3
4 4 4 4
5 5 5 5 5 
ITDS 1372 Computer Programming Date: March 5, 2025

```{r}
for(i in 1:5){
  for(j in 1:i){
    cat(i, " ")
  }
  cat("\n")
}
```

4. Prime Numbers in a Range (1 to 50) Using Nested Loops
• Use a for loop to iterate through numbers 1 to 50.
• Use another for loop to check if the number is prime.
• Print all prime numbers.

```{r}
for (num in 1:50) {
  is_prime <- TRUE  
  if (num < 2) {
    is_prime <- FALSE
  } else {
    for (i in 2:(num - 1)) { 
      if (num %% i == 0) {
        is_prime <- FALSE
      }
    }
  }
  if (is_prime) {
    cat(num, " ")
  }
}
```

5. Print a Diamond Pattern
• Use nested for loops.
• Print a diamond pattern of * with a given height.
ifelse() Problems

```{r}

height <- 5

for (i in 1:height) {
  for (j in 1:(height - i)) {
    cat(" ")
  }
  for (j in 1:i) {
    cat("* ")
  }
  cat("\n")
}

for (i in (height-1):1) {
  for (j in 1:(height - i)) {
    cat(" ")
  }
  for (j in 1:i) {
    cat("* ")
  }
  cat("\n")
}
  
```

6. Assign Pass/Fail Using ifelse()
• Create a vector of student marks.
• Use ifelse() to assign "Pass" if marks are >=50, otherwise "Fail".

```{r}
student_marks <- c(87,90,45,75,62)
ifelse (test=student_marks>=50, yes="Pass",no="Fail")
student_marks
```

7. Check Even or Odd Numbers Using ifelse()
• Create a vector of numbers.
• Use ifelse() to return "Even" for even numbers and "Odd" for odd numbers.

```{r}
numbers <- c(100,75,35,78,91,1001)
ifelse (test=numbers %% 2== 0, yes="Even", no="Odd")
```

8. Assign Grades Based on Marks
• Create a vector of student scores.
• Use ifelse() to assign grades:
o "A" for >=90,
o "B" for 80-89,
o "C" for 70-79,
o "D" for 60-69,
o "F" for <60.

```{r}
student_scores <- c(95,84,76,54,66,78)
ifelse(test=student_scores>=90, yes="A", ifelse(test=student_scores>=80, yes="B", ifelse(test=student_scores>=70, yes="C", ifelse(test=student_scores>=60, yes="D", no="F"))))
```

9. Replace Negative Values in a Vector with Zero
ITDS 1372 Computer Programming Date: March 5, 2025
• Create a vector with positive and negative values.
• Use ifelse() to replace all negative values with 0.

```{r}
vector <- c(-60, 70, 103, -40, -100)
ifelse (test=vector>0, yes=vector, no=0)
```

10. Categorize Age Groups Using ifelse()
• Create a vector of ages.
• Use ifelse() to categorize:
o "Child" for <18,
o "Adult" for 18-60,
o "Senior" for >60.

```{r}
ages <- c(6,45,36,21,57,10,68,53)
ifelse(test=ages>60,yes="Senior", ifelse(test=ages>=18, yes="Adult", no="Child"))
```

Basic Function Problems
1. Write a function to add two numbers
• Create a function add_numbers(a, b) that takes two numbers as input and returns their
sum.
• Call the function with two sample values.

```{r}
a=3
b=6
addition <- function(a,b) {
  sum <- a+b
  return (sum)
}
addition(a,b)
```

2. Write a function to check if a number is even or odd
• Define a function check_even_odd(n) that takes an integer and returns "Even" if it's
even and "Odd" otherwise.
• Test it with different values.

```{r}

even_odd <- function(a) {
  if (a%%2==0){
    return ("Even")
  } else {return ("Odd")}
}
even_odd(10)
```

3. Write a function to calculate the factorial of a number
• Create a function factorial(n) that calculates the factorial using recursion.
• Test the function with n = 5.

```{r}

factorial <- function(a){
b<-1
  for (i in 1:a){
    b<-b*i
  }
  return (b)
}

factorial(5)
```

4. Write a function to find the maximum of three numbers
• Define a function max_of_three(a, b, c) that returns the largest of the three numbers.
• Test with different inputs.
ITDS 1372 Computer Programming Date: March 5, 2025

```{r}
max_of_three <- function (a,b,c) {
  vector <- c(a,b,c)
  maximum <- max(vector)
  return (maximum)
}
max_of_three (1,2,3)
```

5. Write a function to generate Fibonacci sequence up to n terms
• Create a function fibonacci(n) that returns the first n Fibonacci numbers as a vector.
• Example: fibonacci(6) should return [1, 1, 2, 3, 5, 8].

```{r}
fibonacci <- function (a) {
  vector_fib <- c(0,1)
  for (i in 3:(a+1)){
    fib <- vector_fib[(i-1)]+vector_fib[(i-2)]
    vector_fib <- c(vector_fib, fib)
  } 
  return (vector_fib [-1])
}
fibonacci (6)
```

Intermediate Function Problems
6. Write a function to count the number of vowels in a string
• Create a function count_vowels(text) that takes a string and counts the vowels (a, e,
i, o, u).
• Example: count_vowels("hello world") should return 3.

```{r}
count_vowels <- function(text) {
  text <- tolower(text)
  vowels <- c("a", "e", "i", "o", "u")
  char_vector <- unlist(strsplit(text, ""))
  vowel_count <- sum(char_vector %in% vowels)
  return(vowel_count)
}
count_vowels("What's' up?")
```

7. Write a function to reverse a string
• Define a function reverse_string(str) that reverses a given string.
• Example: reverse_string("R programming") should return "gnimmargorp R".

```{r}
reverse_string <- function(str) {
  reversed_str <- paste(rev(strsplit(str,NULL)[[1]]), collapse = "")
  
  return(reversed_str)
}
reverse_string("ABCDE")
```

8. Write a function that accepts a vector and returns only the even numbers
• Create a function filter_even(vec) that takes a numeric vector and returns a new
vector with only even numbers.
• Example: filter_even(c(1,2,3,4,5,6)) should return [2,4,6].

```{r}
filter_even <- function(vec) {
  even_numbers <- vec[vec %% 2 == 0]
  return(even_numbers)
}
filter_even(c(1, 2, 3, 4, 5, 6))
```

Advanced Function Problems
9. Write a function that checks if a given number is prime
• Define a function is_prime(n) that returns TRUE if n is a prime number and FALSE
otherwise.
• Test it with multiple numbers.
ITDS 1372 Computer Programming Date: March 5, 2025

```{r}
is_prime<- function (num){
  is_prime <- TRUE  
  if (num < 2) {
    is_prime <- FALSE
  } else {
    for (i in 2:(num - 1)) { 
      if (num %% i == 0) {
        is_prime <- FALSE
      }
    }
  }
  return (is_prime)
}
is_prime(31)
```

10. Write a function that calculates the mean and standard deviation of a numeric vector
• Create a function stats_summary(vec) that returns a list containing both the mean and
standard deviation of vec.
• Example: stats_summary(c(10, 20, 30, 40, 50)) should return a list like:
$mean
[1] 30
$sd
[1] 15.81

```{r}
stats_summary <- function (vec){
  stdev <- sd(vec, na.rm = FALSE)
  mean <- mean(vec, na.rm = FALSE)
  result <- list(mean=mean,sd=stdev)
  return(result)
}
stats_summary(c(1,2,3,4,5,6,7,NA))
```


Functions: Using ifelse()

1. Categorize Numbers as Positive, Negative, or Zero
categorize_numbers <- function(vec) {
 return(ifelse(vec > 0, "Positive", ifelse(vec < 0, "Negative", "Zero")))
}
# Test the function
numbers <- c(-5, 0, 10, -3, 8, 0)
categorize_numbers(numbers)
Output:
[1] "Negative" "Zero" "Positive" "Negative" "Positive" "Zero"

```{r}
categorize_numbers <- function(vec) {
 return(ifelse(vec > 0, "Positive", ifelse(vec < 0, "Negative", "Zero")))
}
# Test the function
numbers <- c(-5, 0, 10, -3, 8, 0)
categorize_numbers(numbers)
```


2. Assign Grades Based on Scores
assign_grades <- function(scores) {
 return(ifelse(scores >= 90, "A",
 ifelse(scores >= 80, "B",
 ifelse(scores >= 70, "C",
 ifelse(scores >= 60, "D", "F")))))
}
# Test the function
marks <- c(95, 82, 74, 60, 45)
assign_grades(marks)
Output:
[1] "A" "B" "C" "D" "F"
ITDS 1372 Computer Programming Date: March 5, 2025

```{r}
assign_grades <- function(scores) {
 return(ifelse(scores >= 90, "A",
 ifelse(scores >= 80, "B",
 ifelse(scores >= 70, "C",
 ifelse(scores >= 60, "D", "F")))))
}
# Test the function
marks <- c(95, 82, 74, 60, 45)
assign_grades(marks)
```

3. Categorize Temperature Levels
categorize_temperature <- function(temp) {
 return(ifelse(temp >= 30, "Hot",
 ifelse(temp >= 20, "Warm",
 ifelse(temp >= 10, "Cool", "Cold"))))
}
# Test the function
temps <- c(35, 25, 15, 5, 28)
categorize_temperature(temps)
Output:
[1] "Hot" "Warm" "Cool" "Cold" "Warm"

```{r}
categorize_temperature <- function(temp) {
 return(ifelse(temp >= 30, "Hot",
 ifelse(temp >= 20, "Warm",
 ifelse(temp >= 10, "Cool", "Cold"))))
}
# Test the function
temps <- c(35, 25, 15, 5, 28)
categorize_temperature(temps)
```

4. Check Age Group (Child, Teen, Adult, Senior)
age_group <- function(ages) {
 return(ifelse(ages < 13, "Child",
 ifelse(ages < 20, "Teen",
 ifelse(ages < 60, "Adult", "Senior"))))
}
# Test the function
ages <- c(10, 16, 25, 45, 70)
age_group(ages)
Output:
[1] "Child" "Teen" "Adult" "Adult" "Senior"

```{r}
age_group <- function(ages) {
 return(ifelse(ages < 13, "Child",
 ifelse(ages < 20, "Teen",
 ifelse(ages < 60, "Adult", "Senior"))))
}
# Test the function
ages <- c(10, 16, 25, 45, 70)
age_group(ages)
```

5. Classify Numbers as Small, Medium, or Large
classify_numbers <- function(vec) {
 return(ifelse(vec < 10, "Small",
 ifelse(vec < 50, "Medium", "Large")))
}
# Test the function
values <- c(5, 20, 45, 75, 100)
classify_numbers(values)

```{r}
classify_numbers <- function(vec) {
 return(ifelse(vec < 10, "Small",
 ifelse(vec < 50, "Medium", "Large")))
}
# Test the function
values <- c(5, 20, 45, 75, 100)
classify_numbers(values)

```

