---
title: "DataTrans_DatTidy_Lab"
author: "Andie Anger"
date: "3/31/25"
output: html_document
---

###Data Transformation

##Start by loading library(tidyverse) and library(dplyr) if not done already.

```{r}
library(dplyr)
library(tidyverse)
library(nycflights13)
library(tibble)
install.packages("nycflights13")
install.packages("dplyr")
install.packages("tidyverse")
install.packages("tibble")


```


3.2.5 Exercises
In a single pipeline for each condition, find all flights that meet the condition:

1. Had an arrival delay of two or more hours

```{r}
flights |>
  filter(arr_delay >= 120)
```

2. Flew to Houston (IAH or HOU)

```{r}
flights |>
  filter(dest == "IAH" | dest=="HOU") 
```

3. Were operated by United, American, or Delta

```{r}
flights|>
  filter(carrier=="DL" | carrier=="AA" | carrier=="UA") 
```

4. Departed in summer (July, August, and September)

```{r}
flights|>
  filter(month %in% c(7:9)) 
```

5. Arrived more than two hours late but didn’t leave late

```{r}
flights|>
  filter(dep_delay<=0 & arr_delay>120) 
```

6. Were delayed by at least an hour, but made up over 30 minutes in flight

```{r}
flights|>
  filter(dep_delay >= 60 & arr_delay < dep_delay - 30) 
```

7. Sort flights to find the flights with the longest departure delays. Find the flights that left earliest in the morning.

```{r}
flights|>
  arrange(desc(dep_delay),sched_dep_time)
```


8. Sort flights to find the fastest flights. (Hint: Try including a math calculation inside of your function.)

```{r}
flights |> 
  mutate(
    speed = distance / air_time * 60
  ) |> arrange(speed)
```

9. Was there a flight on every day of 2013?

```{r}
all_dates_2013 <- tibble(date = seq.Date(from = as.Date("2013-01-01"), 
                                         to = as.Date("2013-12-31"), by = "day"))

# Use a pipeline to check for flights every day in 2013
missing_dates <- flights |> 
  filter(year == 2013) |>  # Filter for flights from 2013
  select(year, month, day) |>  # Select year, month, and day
  distinct() |>  # Get unique combinations of year, month, and day
  mutate(date = make_date(year, month, day)) |>  # Create a date column
  select(date) |>  # Only keep the date column
  anti_join(all_dates_2013, by = "date")  # Check if any dates are missing

# Display the results
if (nrow(missing_dates) == 0) {
  print("There was a flight every day of 2013.")
} else {
  print("There were days with no flights.")
  print(missing_dates)  # Show the dates with no flights
}
```


10. Which flights traveled the farthest distance? Which traveled the least distance? 

```{r}
flights |> 
  arrange(desc(distance)) |>
  distinct(distance, .keep_all=TRUE) |>
  slice_head(n = 5)


flights |>
  arrange(distance) |>
  distinct(distance, .keep_all=TRUE) |>
  slice_head(n = 5)
```


11. Does it matter what order you used filter() and arrange() if you’re using both? Why/why not? Think about the results and how much work the functions would have to do. 
#It does not change the result by changing the order of the filter() and arrange() functions.

###Displaying three tables

TB cases in world

###5.2.1 Exercises
For each of the sample tables, describe what each observation and each column represents.

Sketch out the process you’d use to calculate the rate for table2 and table3. You will need to perform four operations:

1.  Extract the number of TB cases per country per year.
2.  Extract the matching population per country per year.
3.  Divide cases by population, and multiply by 10000.
4.  Store back in the appropriate place.



You haven’t yet learned all the functions you’d need to actually perform these operations, but you should still be able to think through the transformations you’d need.


```{r}
table1
```

#To convert to table 2:

```{r}
pivot_longer(table1, cols=3:4, names_to = "type", values_to = "count", values_drop_na = FALSE) 
```

```{r}
table2
```

#To convert to table 3:
```{r}
mutate(table1,rate=cases/population*10000,.keep="unused")
```


```{r}
table3
```
```{r}
# Compute rate per 10,000
table1 |>
  mutate(rate = cases / population * 10000)


# Compute total cases per year
table1 |> 
  group_by(year) |> 
  summarize(total_cases = sum(cases))


# Visualize changes over time
ggplot(table1, aes(x = year, y = cases)) +
  geom_line(aes(group = country), color = "grey50") +
  geom_point(aes(color = country, shape = country)) +
  scale_x_continuous(breaks = c(1999, 2000)) # x-axis breaks at 1999 and 2000
```
