---
title: 'Lab2: Matrix, Lists, and Data Frames'
author: "Andrea Anger"
date: "` 2/28/25"
output: html_document
---
1. Create a Matrix
mat <- matrix(1:9, nrow = 3, ncol = 3)
print(mat)
Creates a 3×3 matrix filled with numbers from 1 to 9.

```{r}
mat <- matrix(1:9, nrow = 3, ncol = 3)
print(mat)


```
 
2. Access Elements
mat[2, 3]  # Access the element in row 2, column 3
mat[1, ]   # Access the first row
mat[, 2]   # Access the second column
 
```{r}
mat[2, 3]  # Access the element in row 2, column 3
mat[1, ]   # Access the first row
mat[, 2]   # Access the second column
```


3. Transpose a Matrix
t_mat <- t(mat)
print(t_mat)
Swaps rows and columns.
```{r}
t_mat <- t(mat)
print(t_mat)
```
 
4. Matrix Addition and Subtraction
mat2 <- matrix(9:1, nrow = 3, ncol = 3)
sum_mat <- mat + mat2
diff_mat <- mat - mat2
Adds and subtracts corresponding elements.
```{r}
mat2 <- matrix(9:1, nrow = 3, ncol = 3)
sum_mat <- mat + mat2
diff_mat <- mat - mat2


```

 
5. Matrix Multiplication
prod_mat <- mat %*% mat2
Performs matrix multiplication.

```{r}

prod_mat <- mat %*% mat2


```

 
6. Element-wise Multiplication
elementwise_prod <- mat * mat2
Multiplies corresponding elements.

```{r}

elementwise_prod <- mat * mat2



```

 
7. Find Determinant
det(mat)
Computes the determinant of a square matrix.

```{r}
det(mat)



```

 
8. Find Inverse of a Matrix
inv_mat <- solve(mat)
Finds the inverse if the matrix is invertible.
```{r}

inv_mat <- solve(mat)


```
 
 
 
9. Binding Matrices
row_bind <- rbind(mat, c(10, 11, 12))  # Adds a row
col_bind <- cbind(mat, c(10, 11, 12))  # Adds a column

```{r}

row_bind <- rbind(mat, c(10, 11, 12))  # Adds a row
col_bind <- cbind(mat, c(10, 11, 12))  # Adds a column

```

 
10. Checking Matrix Dimensions

dim(mat)
nrow(mat)
ncol(mat)
Returns the number of rows and columns in a matrix.

```{r}
dim(mat)
nrow(mat)
ncol(mat)


```

 
11. Extracting a Submatrix


sub_mat <- mat[1:2, 2:3]
print(sub_mat)
Extracts a 2×2 submatrix from the original matrix.
```{r}

sub_mat <- mat[1:2, 2:3]
print(sub_mat)


```


1. Creating a List

my_list <- list(name = "Alice", age = 25, scores = c(85, 90, 95))
print(my_list)
Creates a list with different data types.

```{r}
my_list <- list(name = "Alice", age = 25, scores = c(85, 90, 95))
print(my_list)
```


2. Accessing List Elements by Index

my_list[[1]]  # Access the first element
my_list[[3]]  # Access the third element
Uses [[ ]] to retrieve elements.

```{r}
my_list[[1]]  # Access the first element
my_list[[3]]  # Access the third element

```


3. Accessing List Elements by Name


my_list$name
my_list$scores
Uses $ to access named elements.

```{r}
my_list$name
my_list$scores


```


4. Modifying List Elements

my_list$age <- 26
my_list$scores[1] <- 88  

# Modify an element within a vector inside a list

```{r}

my_list$age <- 26
my_list$scores[1] <- 88  

```

5. Adding a New Element to a List

my_list$city <- "New York"

#Adds a new element to the list.
```{r}
my_list$city <- "New York"


```


6. Removing an Element from a List

my_list$age <- NULL
Removes the age element from the list.
```{r}

my_list$age <- NULL

```



7. Checking List Structure

str(my_list)
Displays the structure of the list.

```{r}
str(my_list)


```


8. Finding the Length of a List

length(my_list)
Returns the number of elements in the list.

```{r}


length(my_list)

```


9. Creating a Nested List

nested_list <- list(person = my_list, company = "TechCorp")
print(nested_list)
A list inside another list.

```{r}

nested_list <- list(person = my_list, company = "TechCorp")
print(nested_list)


```




10. Flattening a Nested List

unlist(my_list)
Converts the list into a vector (if possible).

```{r}


unlist(my_list)

```

1. Creating a Data Frame

df <- data.frame(Name = c("Alice", "Bob", "Charlie"),
                 Age = c(25, 30, 35),
                 Score = c(85, 90, 95))
print(df)
Creates a simple data frame with three columns.
```{r}
df <- data.frame(Name = c("Alice", "Bob", "Charlie"),
                 Age = c(25, 30, 35),
                 Score = c(85, 90, 95))
```
 
2. Accessing Columns by Name

df$Age  # Access the "Age" column
df["Score"]  # Also accesses the "Score" column

```{r}
df$Age  # Access the "Age" column
df["Score"]  # Also accesses the "Score" column
```
 
3. Accessing Rows and Elements by Index


df[1, ]   # First row
df[, 2]   # Second column
df[2, 3]  # Value at row 2, column 3
```{r}

df[1, ]   # First row
df[, 2]   # Second column
df[2, 3]  # Value at row 2, column 3

```
 
4. Adding a New Column


df$City <- c("New York", "Los Angeles", "Chicago")
print(df)
Adds a new column to the data frame.

```{r}
df$City <- c("New York", "Los Angeles", "Chicago")
print(df)
```

 
5. Removing a Column
df$Score <- NULL
Removes the "Score" column.

```{r}
df$Score <- NULL


```

 
6. Filtering Rows Based on Condition


subset(df, Age > 28)
Filters rows where Age is greater than 28.
```{r}

subset(df, Age > 28)


```
 
7. Sorting a Data Frame


df_sorted <- df[order(df$Age), ]
print(df_sorted)
Sorts the data frame by Age in ascending order.
```{r}

df_sorted <- df[order(df$Age), ]
print(df_sorted)

```
 
8. Selecting Multiple Columns


df_selected <- df[, c("Name", "Age")]
print(df_selected)
Extracts only the "Name" and "Age" columns.
 
```{r}
df_selected <- df[, c("Name", "Age")]
print(df_selected)


```
 
9. Finding the Number of Rows and Columns

nrow(df)  # Number of rows
ncol(df)  # Number of columns
dim(df)   # Dimensions of the data frame

```{r}

nrow(df)  # Number of rows
ncol(df)  # Number of columns
dim(df)   # Dimensions of the data frame

```
 
 
10. Checking the Structure of a Data Frame


str(df)
Displays structure including column types and sample values.
 
```{r}

str(df)

```
 
11. Converting a Column’s Data Type


df$Age <- as.character(df$Age)
Changes the "Age" column from numeric to character.
 
```{r}


df$Age <- as.character(df$Age)


```
 
12. Applying a Function to Each Column


colMeans(df[, sapply(df, is.numeric)], na.rm = TRUE)
Finds the mean of all numeric columns.
```{r}


colMeans(df[, sapply(df, is.numeric)], na.rm = TRUE)

```
 
13. Merging Two Data Frames


df2 <- data.frame(Name = c("Alice", "Bob", "Charlie"),
                  Salary = c(50000, 60000, 70000))
merged_df <- merge(df, df2, by = "Name")
print(merged_df)
Merges two data frames based on the "Name" column.

```{r}


df2 <- data.frame(Name = c("Alice", "Bob", "Charlie"),
                  Salary = c(50000, 60000, 70000))
merged_df <- merge(df, df2, by = "Name")
print(merged_df)

```
 
14. Handling Missing Values


df$Age[2] <- NA  # Introduce a missing value
df_cleaned <- na.omit(df)  # Remove rows with missing values
print(df_cleaned)
Handles missing values by removing rows that contain NA.
```{r}

df$Age[2] <- NA  # Introduce a missing value
df_cleaned <- na.omit(df)  # Remove rows with missing values
print(df_cleaned)


```
 
15. Converting a Data Frame to a Matrix


df_matrix <- as.matrix(df)
print(df_matrix)
Converts a data frame into a matrix.

```{r}

df_matrix <- as.matrix(df)
print(df_matrix)


```
16. Creating a Data Frame with stringsAsFactors = TRUE

df1 <- data.frame(Name = c("Alice", "Bob", "Charlie"), 
                  Age = c(25, 30, 35), 
                  stringsAsFactors = TRUE)
str(df1)
The "Name" column is converted into a factor.

```{r}

df1 <- data.frame(Name = c("Alice", "Bob", "Charlie"), 
                  Age = c(25, 30, 35), 
                  stringsAsFactors = TRUE)
str(df1)

```

17. Creating a Data Frame with stringsAsFactors = FALSE (Default in R 4.0)

df2 <- data.frame(Name = c("Alice", "Bob", "Charlie"), 
                  Age = c(25, 30, 35), 
                  stringsAsFactors = FALSE)
str(df2)
The "Name" column remains as a character vector.
```{r}
df2 <- data.frame(Name = c("Alice", "Bob", "Charlie"), 
                  Age = c(25, 30, 35), 
                  stringsAsFactors = FALSE)
str(df2)
```

18. Converting a Character Column to Factor Manually

df2$Name <- as.factor(df2$Name)
str(df2)
Manually converts the "Name" column to a factor.
```{r}
df2$Name <- as.factor(df2$Name)
str(df2)
```

19. Checking if a Column is a Factor

is.factor(df1$Name)  # TRUE because stringsAsFactors = TRUE
is.factor(df2$Name)  # FALSE because stringsAsFactors = FALSE
Verifies whether a column is a factor or not.
```{r}
is.factor(df1$Name)  # TRUE because stringsAsFactors = TRUE
is.factor(df2$Name)  # FALSE because stringsAsFactors = FALSE
```

20. Changing Factor Levels After Using stringsAsFactors = TRUE

levels(df1$Name) <- c("Alicia", "Robert", "Charles")
print(df1)
Renames factor levels in the "Name" column.

```{r}
levels(df1$Name) <- c("Alicia", "Robert", "Charles")
print(df1)
```

